Astrophysical calibration of gravitational-wave detectors

### 2015-04-07 docs/cosmic_cal.tex
Started filling out the paper draft from Liam's report.

@@

\newcommand{\dcc}{LIGO-PXXXXXXXX}
\newcommand{\CM}[1]{\textcolor{red}{CM: #1}}
\newcommand{\MH}[1]{\textcolor{blue}{MH: #1}}
\newcommand{\XF}[1]{\textcolor{green}{XF: #1}}

@@

\title{Requirements and usefulness of galaxy catalogues for gravitational wave astrophysics}

@@

\begin{abstract}
  We present an investigation into the potential of calibrating
  gravitational wave detector outputs through the use of standard
  sirens. Such signals, as measured via gravitational wave
  observations provide an estimated luminosity distance which is
  subject to uncertainties in the calibration of the data.  If a host
  galaxy is identified for a given source then its redshift can be
  obtained and combined with current knowledge of the cosmologic
  parameters.  This will yield the true luminosity distance and allow
  the direct comparison with the estimated value.  Discrepancies can
  then be used to correct the originbal calibration.
\end{abstract}

@@

Explain the basic idea.

@@

We need to define our model for the calibration, i.e. are we talking
about amplitude only and are we breaking this up over frequency.

@@

We should discuss the key points with regards to

@@

This should be the simplest case we can do (and maybe the only case).

@@

An analysis or discussion of the issues regarding multiple events
without EM counterparts.

@@

What did we learn.

>>>>

\newcommand{\curlH}{\mathcal{H}}
\newcommand{\gws}{\tilde{h}}

@@

\newcommand{\dcc}{LIGO-PXXXXXXXX}
\newcommand{\cm}[1]{\textcolor{red}{CM: #1}}
\newcommand{\MP}[1]{\textcolor{blue}{MP: #1}}
\newcommand{\lw}[1]{\textcolor{green}{LW: #1}}

@@

\title{Astrophysical calibration of gravitational-wave detectors}

@@

\begin{abstract}
  We present an investigation into the potential of calibrating
  gravitational wave detector outputs through the use of standard
  sirens. Such signals, as measured via gravitational wave
  observations provide an estimated luminosity distance which is
  subject to uncertainties in the calibration of the data.  If a host
  galaxy is identified for a given source then its redshift can be
  obtained and combined with current knowledge of the cosmologic
  parameters.  This will yield the true luminosity distance and allow
  the direct comparison with the estimated value.  Discrepancies can
  then be used to correct the original calibration.
\end{abstract}

@@

\cm{Explain the basic idea.}

@@

\cm{We need to define our model for the calibration, i.e. are we talking
about amplitude only and are we breaking this up over frequency.}

The technique used throughout GW research in calibrating the detectors is
carried out through a complicated system of physical manipulations of the
Fabret-Perot Michelson interferometer; where an elaborate feedback system is
used to sustain a defined measurement in arm length difference between the
moving mirrors~\cite{LIGOCal}. In general, the interferometers are set to move
at a frequency corresponding to a GW signal. An error signal is then found
through the Pound-Driver-Hall technique~\cite{Black} for mirror stabilisation
which is proportional to the difference in mirror separation of the
Fabret-Perot Michelson interferometer. The Pound-Drever-Hall technique is
essentially a control loop feedback system in which the difference output (in
GW detectors, this is the distance the mirrors move due to the GW) and the the
output set is reduced to the highest degree of certainty. This is just a brief
overview of the calibration method currently used, and is an extensive study in
itself, see~\cite{Vitale} and references therein. For advanced LIGO and Virgo,
at the frequency range used in this project, the error in the calibration using
the 'hardware injection method' is roughly $10\%$~\cite{Vitale}. This is a
benchmark set for the error estimation using the proposed new method in this
project.

The hardware method of calibrating the ground-based detectors is incredibly
complex in its application, with many sources of error which have to be
accounted for. The new method of calibration proposed and tested in this
project is a lot simpler. The method acts as an external calibrator, to be
carried out using the data collected by the detector, $\tilde{d}(f)$. We assume
the collected data contains some unknown and unaccounted for calibration error.
The uncertainty in $\tilde{d}(f)$ can then be evaluated using methods of
parameterisation already carried out in GW research. This method of
parameterising GW defining variables is then carried out on a known source, a
standard candle as described above, which allows one to measure the calibration
uncertainty in the data set.

One can estimate the error in the calibration using parameter estimation
methods, or more precisely Bayesian inference. In this project the calibration
offset used throughout is simply a scale factor, $s$. This effects the data set
collected by a simple multiplication:

\begin{equation}
  \label{eq:scaledata}
  \tilde{d}_m(f) = \tilde{d}_e(f) \times s
\end{equation}

where $\tilde{d}_m(f)$ is the data-set measured by the detector, and
$\tilde{d}_e(f)$ is the exact data set, with no error. A description of the
collected data is given in section~\ref{sec:cbc}. The scale factor being
parameterised, $s$, represents all known sources of error in the system. This
is not entirely realistic as $s$ is known to be a function of frequency.
Testing a more realistic calibration error is work to be done in the future.

Being able to resolve the calibration scale factor with the least amount of
uncertainty was the aim of this project; to show that this method is
competitive with current techniques. If the calibration error can be resolved
with known uncertainty, GW data analysis can then be carried out with more
certainty. This is useful as an entirely independent method of assessing the
uncertainty in the current calibration method.

@@

A GW detection on its own right would be ground breaking, allowing
astrophysicists to observe the sky above in a whole new spectrum. However, to
increase the scientific return on a direct GW detection, a counterpart EM
observation of the source itself would need to be seen.  There are two
approaches to observing a source both in EM and GW spectrum's~\cite{grb}:

\begin{enumerate}
  \item locating and observing a source with an EM follow-up to
a GW detection
  \item searching for a GW signal generated by a source seen
through EM observations
\end{enumerate}

Both of these methods have their pros and cons. Firstly, an EM follow-up would
be very difficult due to the duration of GW source system detectable by
ground-based interferometers. A neutron star binary system (NS-NS binary) has a
duration of $\sim15$ minutes~\cite{bnstime}.  Errors in the sky position from a
GW detection would also create issues for an EM follow-up. The sky localisation
obtained from the ground-based detectors gives a large field of view for the EM
observatories to search, roughly an area spanning $100$'s of
degrees~\cite{grb}.

The more promising detection approach then is a follow-up GW search from an EM
trigger, especially for the source of GW considered in this paper. The merger
time for NS-NS binaries are known with an uncertainty within a few seconds for
the EM radiation emitted from this system~\cite{grb}.

Identifying the source of GWs redshift would then lead to calculating the
Hubble constant and other cosmological parameters~\cite{Schutz}. The evolution
of the system could be better understood, allowing for searches with greater
accuracy. Counterpart EM observations of a GW source will also remove
uncertainty in the defining GW parameters, a fact which was taken advantage of
for this project.

The source of GW considered in this paper will be a coalescing binary system
consisting of neutron stars. As well as being the most probable source of GWs
detectable by ground-based interferometers, they are also expected to emit EM
radiation. As shown in figure~\ref{fig:grb}, the emission occurs due to the
high velocity with which the stars coalesce. The coalescences are thought to be
the progenerators of short-duration gamma ray bursts (SGRBs)~\cite{grb}.
Observations of this radiation has been see, through the Swift satellite and
follow-up ground-based telescopes~\cite{swift}. Unfortunately, no observation
of SGRBs have been seen within the range of advanced generation of LIGO and
Virgo. It is probable, however, that such a source exists well within the range
of advanced LIGO/ Virgo, but has simply not been observed~\cite{grb}. This then
validates the assumption made throughout the project.

In this paper, a key assumption has to be made in order to test this new method
of calibration. It is assumed that there exists a source of GWs with enough
information known through EM observation such that it can be used as a
\textit{standard candle}, i.e. an accurately known and well modelled source of
GWs. In this project (for the source of GWs being analysed), a GRB observation
is assumed to have been seen. This allows for discrepancies in the detector
calibration itself to be parameterised. This is the proposed new method of
calibration for the detectors, for all types of detectors.

@@

\cm{This should be the simplest case we can do.}

In this project the hypothesis considered, $\curlH$, simply states that the
data, $\tilde{d}(f),$ collected by a generic detector contains the GW signal,
$\gws(f)$, and some noise, $\tilde{n}(f)$~\cite{JVei},

\begin{equation}
  \label{eq:data}
  \tilde{d}^D(f) = \gws^D(f) + \tilde{n}^D(f)
\end{equation}

where $D$ represents the detector site. In this project there will be three
detectors in question: LIGO Livingston (L), LIGO Hanford (H), and Virgo (V); $D
= \{L,H,V\}$. Testing different hypotheses is the method used to determine
whether a data set contains a GW signal or not~\cite{SathSchutz}. This was not
explored in this project so the above hypothesis is consistent throughout this
paper.

Coalescing binary systems of compact objects, such as neutron stars or black
holes, are the most probable source of GWs detectable by ground-based
interferometers~\cite{peadvanced}. A direct detection of GWs would allow for a
fuller understanding of the mechanics of these systems. Compact binary
coalescence's (CBCs) are systems of two compact object inspiralling, with
reducing separation, until the inevitable merger of the objects. This orbital
motion creates ripples in space-time; this is the GW from this motion. For
gravitational wave research, the interests lie in the GW signal generated, or
the \textit{waveform}, shown in figure~\ref{fig:signal}.

In this paper, a brief overview of the mathematical breakdown for the waveform
is shown. For a more rigorous derivation, see~\cite{SathSchutz} and references
there in. The waveform used throughout this project is a simplified system
where the neutron stars are non rotating about their axes, in a perfectly
circular inspiralling orbit; known as the TaylorF2 model~\cite{TF2,pN}. A more
complex and realistic model could be tested in future for this calibration
method. Parameter estimation using a variety of complex waveforms are carried
out in~\cite{peadvanced}.

The general equation for a GW signal detected at a generic detector, in the
reference frame of the detector site, is given as~\cite{SathSchutz}

\begin{equation}
  \label{eq:gravsig}
   \gws^D(f) = \big[ F_{+}^D(\alpha, \delta, \psi)\gws_{+}(f) +
F_{\times}^D(\alpha, \delta, \psi)\gws_{\times}(f)\big]
\end{equation}

where $F_{+,\times}(\alpha, \delta, \psi)$ are the antenna response function of
the detector, dependent on the right ascension, $\alpha$, declination,
$\delta$, the polarisation angle $\psi$, and time. Testing the antenna patterns
is carried out in section~\ref{sec:rndsp} of this report when the sky
positions, $(\alpha, \delta)$, are randomised across the entire field of view
for the detector.

The polarisation amplitudes of the GW, given by $\gws_{+}(f)$, and
$\gws_{\times}(f)$, represent the direction with which the GW compresses and
stretches space-time. The form of these two amplitudes depends on the source
being analysed. As mentioned above, the model used throughout this project was
the TaylorF2 model. The parameters defining the GW polarisation amplitudes
using the TaylorF2 model are:

\begin{itemize}
\item the total mass, $M = m_{1}+m_{2}$, where $m_{1,2}$ are the masses of the
binary stars
\item the symmetric mass ratio $\eta = \frac{m_{1}m_{2}}{(m_{1}+m_{2})}$
\item chirp mass, $\mathcal{M} = \eta^{3/5}M$
\item luminosity distance, $D_{L}$, measured in Mpc ($1 Mpc \approx
3.1\times10^{22}m  $)
\item the angle of inclination $\iota$. This is the angle from the source,
taken as the angle between the normal of the orbital plane and the observer
orientation.
\item phase of the wave, $\psi$, which defines the polarisation of the angle of
the GW, shown in equation~\ref{eq:psi}.
\item the coalescence time, $t_c$, and phase of coalescence, $\phi_c$.
\end{itemize}

Together with the sky position, these parameters form the vector $\vec{\Theta}
= \{\mathcal{M}, \eta, D_{L}, \iota, \alpha, \delta, \psi, t_{c}, \phi_{c}\}$
which defines the entire GW signal.  The form of the polarisation amplitudes
using the TaylorF2 model are given as~\cite{SathSchutz,pN,TF2}

\begin{equation}
  \gws_{+}(f) = \frac{1+\cos^{2}\iota}{2D_{L}}
\frac{\mathcal{M}^{5/6}}{\pi^{2/3}} \sqrt{\frac{5}{24}}f^{-7/6}e^{-i\psi(f)}
\end{equation}
\begin{equation}
  \gws_{\times}(f) =
i\frac{\cos^{2}\iota}{D_{L}}\frac{\mathcal{M}^{5/6}}{\pi^{2/3}}
\sqrt{\frac{5}{24}}f^{-7/6}e^{-i\psi(f)}
\end{equation}

where the phase of the GW, $\psi$, is
\begin{equation}
  \label{eq:psi}
  \psi(f) = 2\pi f t_c - \phi_c + \psi_N(\eta)\sum\limits_{k=0}^4{\psi_k(\eta)
(\pi(m_1 + m_2)f)^{(k-5)/3}}.
\end{equation}

The terms $\psi_N$ and $\psi_k$ are the standard Newtonian and post-Newtonian
coefficients. Derivations of these two equations can be found in~\cite{pN} and
references there-in. The simulated gravitational wave signal used throughout
this project can be seen in figure~\ref{fig:signal}. As one can see, the signal
strength is very faint (an order of magnitude around $10^{-24}$). This is the
the biggest problem in detecting GWs.

As mentioned above, the source considered throughout this project thought to
produce EM radiation; SGRBs. In this project, we assume that SGRBs of the GW
source described above have been observed. If a GRB has been seen, then the
redshift of the merger can be calculated. This then allows for the luminosity
distance to be calculated and in turn other defining parameters can be obtained
(or known to a degree of uncertainty)~\cite{SathSchutz}.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{grav_signal}
  \caption{The gravitational waveform produced using coalescing binary
neutron stars. The wave is represented in the frequency domain with masses
$m_{1,2} = 1.4M_{\odot}$ at a luminosity distance $D_{L} = 200 \,Mpc$.}
  \label{fig:signal}
\end{figure}

%More explination
Finally, the last term in equation~\ref{eq:data} must be discussed; the noise
of the model, $\tilde{n}(f)$. Throughout this project, we have added no noise
to the signal, although as discussed later we do include a noise estimate in
the likelihood.

The prior probability function, $p(\vec{\theta} | \curlH)$ in
equation~\ref{eq:bayesgw}, represents our prior knowledge in $\theta$ before
the analysis.The range of parameters used throughout this paper were chosen due
to their inability to be justifiably known through a GRB assumption. The
parameters were drawn from the ranges for any signals generated in this
analysis, given in table 1.

\begin{table}
  \centering
  \begin{tabular}{|c|c|}
%  \caption{Prior parameters}
    \hline
    Parameter & Distribution \\
    \hline
    $\psi$   &  U($0,\pi/2$)\\
    $\iota$   & N($0,0.1$) \\
    $\phi_c$  & U($0,\pi/2$)\\
    $t_c$     & U($t_{o}-0.01, t_{o}+0.01$) \\
    $s$       & U(0,2) \\
    \hline
  \end{tabular}
  \label{tab:prior}
  \caption{\textit{The parameters used throughout the analysis with some degree
of uncertainty. U(a,b) is a uniform distribution between a and b. N($\mu,
\sigma$) is a normal distribution, with a mean $\mu$ and standard deviation
$\sigma$. $t_{o} = 9\times10^{8}$}}
\end{table}

Through the assumption made above allows us to constrain the system,
maintaining a small number of parameters in the prior. For a realistic
parameter estimation of the source, a larger range of values with a bigger
distribution. This was not tested due to time constraints but is proposed as
work for the future.

The likelihood functiona used throughout this project is given for a generic
detector D as~\cite{JVei},

\begin{equation}
  p(\vec{d} | \vec{\theta}, \curlH, I) \propto \exp(-(d^{D}(f) -
h^{D}(f)|d^{D}(f) - h^{D}(f))/2)
\end{equation}

where $(\cdot |\cdot )$ is a representation of the inner product. This
derivation is given in the appendix. Using the inner product of two real
functions, and approximating for discrete integer values, the likelihood then
becomes

\begin{equation}
  p(\vec{d} | \vec{\theta}, \curlH, I) \propto \exp\Bigg[ 2\Delta
f\sum_{k>0}\frac{|\tilde{d}(f_{k}) - \tilde{h}(f_{k})|^{2}}{S(f_{k})} \Bigg]
\end{equation}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{PSD}
  \caption{\textit{Noise spectral density for advanced LIGO and Virgo.}}
  \label{fig:psd}
\end{figure}

where $S(f_{k})$ is the \textit{power spectral density} (PSD). This defines the
sensitivity of the detector as it a measure of the strain in the detector due
to noise~\cite{SathSchutz,sens}. The PSD for advanced LIGO and Virgo are shown
in figure~\ref{fig:psd}. The figure shows that the strain effect for advanced
LIGO is less than it is for Virgo; LIGO has a better sensitivity. This is a
factor in analysing the calibration error, given in section~\ref{sec:results}.
The subscript $k$ represents discrete values of relevant frequency bins (As
carrying out the integral would become computationally costly). This likelihood
will be used throughout the project. The factor of $\Delta f$ comes from the
approximation of the inner product of two real functions (as it is integrated
over frequency space).

The aim of this project was to determine how well the calibration error could
be parameterised in a software injected gravitational wave source. The error
was estimated through numerous situations but the basis of the simulations
remained the same throughout the project: a single source of GWs at varying
signal to noise ratios. The GRB assumption was analysed as well by relaxing
some of the defining GW paramaters and adding some uncertainty into their
value, given in section~\ref{sec:pandl}.

These values where chosen due to the assumed counterpart EM observation. These
values could not be justified by that assumption so they were added into the
prior with reasonable uncertainty.

The estimation of the error, given in its simplest form as a scale factor, was
performed for numerous systems: single detector, a network of detectors, random
sky positions, etc. Before the parameter estimation was carried out, the
relationship between distance of the source relative to the detector and the
signal to noise ratio (SNR) needed to be found.

The SNR is given by
\begin{equation}
  \label{eq:snr}
  SNR = \sqrt{\Bigg( \frac{2 |\gws^D(f)|^2}{S^D(f)} df \Bigg)}
\end{equation}

where the noise spectral density, $S^D(f)$ is not a function of distance, shown
in section~\ref{sec:pandl}. For coalescing binary neutron stars of masses
$m_{1,2} = 1.4\,M_{\odot}$ and  $D_{L}200\,Mpc$, the expected SNR is roughly 8.
The SNR was determined from $D_{L} = [10,200]\,Mpc$  using a randomly chosen
sky position which defines the antenna patterns and the GW signal. The range of
luminosity distance used in this paper represented a high level of SNR up to
the distance at which the first predicted detectable GW source; $D_{L} =
200\,Mpc$. The detector limits for the advanced era is a luminosity distance of
$\sim 450\,Mpc$. This was not initially tested as the SNR was low enough to
cause problems with computational time. It was however tested in the random sky
position simulations, see section~\ref{sec:rndsp}. The initial value for the
calibration error used in the analysis was chosen randomly.

\begin{figure}
  \centering
  \includegraphics[width = \textwidth]{DvsSNR-mult-det}
  \caption{The distance and SNR relationship for all three detectors:
LIGO Hanford, H1, LIGO Livingston, L1, and VIRGO.}
  \label{fig:distsnr}
\end{figure}

To begin with, parameter estimation was carried out on the simplest case: a
single source of GWs using a single detector, LIGO Hanford, at different levels
of SNR. The aim was to resolve the calibration scale factor with the highest
degree of certainty.  Before testing the uncertainty resolution using a the
full range of prior parameters, we fixed all values to known signal values and
only searched over the scale factor. Results shown for with and without a full
prior range are given in figures~\ref{fig:sd-empty-D}
and~\ref{fig:sd-non-empty-D} respectively.

\begin{figure}
  \centering
  \includegraphics[width = \textwidth]{SD_empty_D10_200}
  \caption{\textit{Posterior distribution on the calibration scale factor for
LIGO Hanford. Simulations carried out using an entirely known source of GWs.}}
  \label{fig:sd-empty-D}
\end{figure}

The figure above is a histogram of the calibration error, with each curve
representing a different distance (SNR). The black line at $s = 0.8$ is the
initial scale factor value used to multiply the unanalysed data set. This
representation of the initial calibration scale factor will be continuous
throughout the paper.  One would expect with lower SNR (greater $D_{L}$), the
estimation of the parameter would become less certain. This is clearly shown in
figure~\ref{fig:sd-empty-D} as the standard deviations of the curves increases
with smaller SNR/larger luminosity distances. For this entirely known source of
GWs, we would expect to find the uncertainty in $s$ to be approximately
$1/$SNR; this is what we see in figure~\ref{fig:sd-empty-D}. The error in
estimating $s$ for $D_{L} = 200 \,Mpc$ is roughly $\sim 20\%$. This is larger
than the current method of obtaining the uncertainty in calibration. A full
test of the new method proposed is carried out in section~\ref{sec:rndsp}.

The next estimation of the calibration scale factor was performed with a full
range of prior parameters, giving a more realistic system. As before, the
simulations were carried out for differing levels of SNR and a histogram of the
calibration error was made, seen in figure~\ref{fig:sd-non-empty-D}.

\begin{figure}
  \centering
  \includegraphics[width = \textwidth]{SD_non_empty_D10_200}
  \caption{\textit{Posterior distribution on the calibration scale factor for
LIGO Hanford. Simulations carried out with some parameters in the prior.}}
  \label{fig:sd-non-empty-D}
\end{figure}

The error in the calibration scale factor, as before, increases with decreasing
SNR (increasing $D_{L}$). The magnitude of this error however is larger than in
the previous simulation. With a full prior range given in
section~\ref{sec:pandl}, the standard deviation increases; at $D_{L} =
200\,Mpc$ the error in $s$ is $\sim 24\%$. When comparing this new method of
calibration scale factor estimation with the current, feedback loop system
method used today, this method is not as accurate. However, it should again be
noted that this method is an independent estimate of the calibration accuracy.

A network of detectors are used throughout GW research. Being able to show the
effect of the calibration error and recovering its uncertainty in the analysis
would help in giving a detection more confidence. In this section, the
simulations were carried out using multiple detectors; LIGO Hanford, LIGO
Livingston, and Virgo. In reality, multiple detectors are used in order to
determine the (rough) sky position of a signal. Due to their locations, if one
site detects a signal before the others, then the position of the source can be
roughly worked out. The aim was, as before, to estimate the calibration scale
factor with the least amount of uncertainty for differing levels of SNR for a
single source of GWs.

For a network of detectors, the overall likelihood function becomes a product
of each likelihood for all detectors, i.e.

\begin{equation}
  \label{eq:mult-likeli}
  p(\vec{d}| \vec{\theta}, \curlH) = \prod \limits_D p(\vec{d}^D| \vec{\theta},
\curlH)
\end{equation}

where $D = \{H,L,V\}$. The initial calibration scale factors where again chosen
at random ($s = [4.6, 2.5, 6.1]$ for LIGO Hanford, LIGO Livingston, and Virgo
respectively). The simulations were the same as before, using the same
parameter values. The GRB assumption was again tested in these simulations as a
handful of parameters were relaxed and given some uncertainty. The results of
these simulations are shown in figure~\ref{fig:mult-empty-D} and
figure~\ref{fig:mult-non-D}.

\begin{figure}
  \centering
  \includegraphics[width = \textwidth]{MD_empty_D10_200}
  \caption{Posterior distribution on the calibration scale factor for
all three detectors. Simulations carried out using an entirely known source of
GWs.}
    \label{fig:mult-empty-D}
\end{figure}

Figure~\ref{fig:mult-empty-D} shows the histograms of the calibration scale
factor for each detector. These simulations used an entirely known source of
GWs, at a randomly chosen sky position (the same sky position as before). As
the figure shows both LIGO sites, Hanford and Livingston, are able to resolve
the error with a measurable degree of uncertainty. The standard deviation for
$s$ at $D_{L}=200\,Mpc$ is $24\%$ and $41\%$ of the true scale factor for
Hanford and Livingston respectively. The estimation for the calibration error
using VIRGO however, was not as accurate. It can be seen that, for increasing
luminosity distances, the resolution of $s$ becomes increasingly uncertain
until the point at which the distribution becomes almost uniform; $D_{L}
\approx 120\,Mpc$. There are a few reasons why this may be. Firstly, the sky
position used may not be 'ideal' for generating the antenna patterns. The most
probable cause of the estimation being very poor will lie in the detector
itself. The interferometer itself is not as sensitive the two LIGO sites, hence
the estimation of parameters using this site will not be as precise.

\begin{figure}
  \centering
  \includegraphics[width = \textwidth]{MD_non_empty_D10_200}
  \caption{Posterior distribution on the calibration scale factor for
all three detectors. Simulations carried out with some prior parameters.}
  \label{fig:mult-non-D}
\end{figure}

The histograms shown in figure~\ref{fig:mult-non-D} were made using the full
range of prior parameters. The uncertainty in the calibration error is larger
with a less known source of GWs. VIRGO, again, has the least certain
calibration error resolution with the distribution becoming even more uniform.
The standard deviation for both LIGO curves become larger with increasing
luminosity distance as before. For $D_{L} = 200Mpc$,  the standard deviation is
$\tilde 30\%$ of the true value. When comparing this to the current method,
this again is not as accurate.

Throughout the duration of this project the antenna patterns were defined by a
randomly chosen sky position, as seen in equation~\ref{eq:gravsig}. This did
not give a true representation of the estimation capable by algorithm as this
position could be a bad one; making parameterisation more difficult and less
precise. To then test this new method of calibration fully, the entire field of
view of the detector must be analysed. The sky positions $\alpha$, $\delta$,
were distributed as: \begin{table}[H] \centering \begin{tabular}{|c|c|} \hline
Parameter & Distribution \\ \hline $\alpha$       & U($0$,$2\pi$) \\
$\sin(\delta)$ & U($-1$,$1$)    \\ \hline \end{tabular} \caption{\textit{The
distribution of sky position. U(a,b) indicates a uniform distribution between a
and b.}} \label{tab:sp} \end{table}

Figures~\ref{fig:snr-rndsp} and~\ref{fig:std-rndsp} were obtained using a
luminosity distance $D_{L} = 200 Mpc$ as this is the expected luminosity
distance predicted to host the first observable GW~\cite{peadvanced}. $100$
different sky positions were tested using the LIGO Hanford site. The number of
locations was limited by the length of time these simulations took to carry
out. Increasing the number of locations would give a more accurate
representation of the estimation ability; this can be carried out in future.
Finally, a full range of prior parameters were also used in the simulation
set-up.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{rand_sp_snr_D200}
  \caption{Signal to noise ratio for every sky position analysed at
$D_{L} = 200 \,Mpc$. The vertical black line represents the SNR for the sky
position used throughout previous simulations}
  \label{fig:snr-rndsp}
\end{figure}

Figure~\ref{fig:snr-rndsp} shows the histogram of SNRs for the entire field of
view for the detector. Predictions state that at a luminosity distance of
$D_{L}=200\,Mpc$, the SNR will be roughly 8. This is consistent with the
results obtained. The vertical black line shown in figure~\ref{fig:snr-rndsp}
represents the SNR used throughout the project using the randomly chosen
$\alpha$ and $\delta$. One can see that this is a less-than-average SNR and may
explain the resolution of the calibration error not being too precise. This
then is a promising result and the full testing of this new calibration method
can be carried out.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{rand_sp_std_D200}
  \caption{Standard Deviation (as a $\%$ of the true scale factor
value) for every sky position analysed at $D_{L}=\,200Mpc$. The vertical black
line indicates the median of the distribution.}
  \label{fig:std-rndsp}
\end{figure}

Figure~\ref{fig:std-rndsp} is the distribution of fractional standard
deviations on s for each sky position analysed, given as a percentage. The
vertical black line in this figure shows the \textit{median} of the
distribution. The results show that the algorithm can resolve the calibration
error in the data within the range of $\sim 7\%$ to $\sim 14\%$ of the true
value of $s$, for $D_{L}=200\,Mpc$. Being able to recover the scale factor
error to this accuracy range confirms that the new method of calibration method
is very competative with current techniques. These results also confirm that
the sky position used throughout the project was indeed a bad choice with an
error estimation of $\sim 24\%$.

Other simulations relating to the randomisation of $\alpha$ and $\delta$ were
carried out in this project. Different luminosity distances were simulated to
test different levels of SNR. The expected trend was that the standard
deviation scales with SNR; with high SNR yielding a better resolution of the
scale factor error. The results show this trend, given in table

\begin{table}
  \centering
  \begin{tabular}{|c|c|c|}
    \hline
    Distance ($Mpc$) &$\langle$ SNR $\rangle$ & median(Standard
Deviation)($\%$) \\
    \hline
    $20$ & 74 & 1.5 \\
    $120$ & 12  & 7 \\
    $450$ & 3 & 27\\
    \hline
  \end{tabular}
  \caption{\textit{Random sky positions for different luminosity distances;
$\langle$SNR$\rangle$ is the mean SNR for the distribution. }}
  \label{tab:dist-randsp}
\end{table}

The last, and lengthiest, set of simulations carried out used a network of
detectors. Analysis on a luminosity distance of $200 \,Mpc$ was the only one
carried out in this paper. This was due to time taken to process these
simulations. With a full prior range and $100$ random sky positions as before,
the standard deviations and SNRs were obtained. Results are shown
in~\ref{tab:mult-rndsp}.

\begin{table}
\centering
\begin{tabular}[H]{|c|c|c|}
\hline
  Detector   & $\langle$SNR$\rangle$ & median(Standard Deviation)($\%$)\\
\hline
LIGO Hanford & 20& 3 \\
LIGO Livingston & 20 & 10 \\
Virgo & 5 & 30\\
\hline
\end{tabular}
\caption{\textit{Results from randomising the sky position using a network of
detectors with a full range of prior parameters at $D_{L} = 200\, Mpc$.}}
\label{tab:mult-rndsp}
\end{table}

Table~\ref{tab:mult-rndsp} gives promising results for both LIGO sites. One can
see that the error estimation for LIGO Hanford is very good, with half of the
positions analysed having an uncertainty $<3\%$. One would expect the gap in
estimation accuracy between the two LIGO sites to be smaller as they are built
to the same sensitivity. This leads to the conclusion that not enough sky
positions were analysed. Virgo again has a less accurate recovery of the
calibration scale factor. As mentioned above, the larger uncertainty is
expected to originate from the sensitivity of the detector itself.

@@

\cm{A discussion of the issues regarding multiple events
without EM counterparts.}

@@

\cm{What did we learn.}

This paper has shown that for differing levels of SNR, the error in the
calibration is resolved within a certain accuracy range. Initial simulations
used a bad sky position, resulting a low SNR which created larger uncertainty
in resolving $s$. It was shown that for this sky position, the calibration
scale factor estimation was recovered with an accuracy of roughly a $27\%$ for
$D_{L} = 200\,Mpc$ for LIGO Hanford. The resolution of $s$ for a network of
detectors was also shown.

The main results obtained in this project were from analysing randomly chosen
sky positions from the entire field of view for the detector. Through these
simulations, it was shown that this new method of calibration can reach and
surpass the current method in terms of accuracy. The results in
figure~\ref{fig:std-rndsp} show that for half of the randomly chosen sky
positions relate to an estimation of $s$ within a range comparable with the
current methods.

These results are reassuring for parameter estimation of the defining GW signal
and would help in reducing the uncertainty in a direct detection. This project
was; however, time constrained. There are many, many things that could be done
with this project. A true noise realisation with each frequency bin having a
random Gaussian noise would be the next step in calibration error estimation.

This project is only the beginning in testing this new method of calibration.
More rigorous testing has to be carried out for a number of different systems.
As mentioned above, a larger number of random sky positions have to be analysed
to give better understanding of the resolution capabilities of the method. The
current technique of calibration defines the error in the detector to be a
complex function of frequency and phase of the GW. Instead of the simple scale
factor used so far, a more realistic error can be tested to ensure that the
estimation in its value is still comparable with current methods.

For applications of this calibration method, hypothesis testing could be
carried out. Testing different hypothesis is the technique used in determining
a GW signal in a data-set when compared to a noise-only data-set. Being able to
show with confidence that the data contains a signal is the primary focus of GW
research at the moment. Therefore if uncertainty can be removed from the
analysis this will ensure a more confident detection. Using this new method is
resolving and removing the uncertainty is an area which has to be explored.
